---
title: "A mixture of a Poisson-lognormal distribution"
author: "Masatoshi Katabuchi"
date: "September 2018"
bibliography: poilog_mix.bib
output:
  #rmdformats::material
  html_document:
    fig_caption: yes
    theme: spacelab #readable #sandstone #spacelab #flatly
    highlight: pygments #tango #kate
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      collapsed: FALSE
      smooth_scroll: TRUE
---

<meta name="twitter:card" content="summary_large_image">
<meta name="twitter:site" content="@mattocci">
<meta name="twitter:creator" content="@mattocci">
<meta name="twitter:title" content="A mixture of a Poisson-lognormal distribution">
<meta name="twitter:description" content="Two-component Possion-lognormal mixtrue">
<meta name="twitter:image" content="/poilog_mix_files/figure-html/mix_dens-1.png">

```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```
<a href="https://twitter.com/mattocci"><i class="fa fa-twitter fa-1x"></i> Twittter</a>
<a href="https://github.com/mattocci27/"><i class="fa fa-github fa-1x"></i> Github</a>
<a href="mailto:mattocci27@gmail.com"><i class="fa fa-envelope fa-1x"></i> Email</a>

Here is a little example of fitting mixture model with R and Stan.

# Mixture model

Given a set of likelihood functions $\pi_1(\boldsymbol y \mid
\alpha_1), \ldots, \pi_k(\boldsymbol y \mid \alpha_K)$ for  variables $\boldsymbol y = (y_1, \ldots, y_N)$, parameterized by $\boldsymbol \alpha = (\alpha_1, \ldots, \alpha_K)$ and mixture weights $\boldsymbol \theta = (\theta_1, \ldots, \theta_K)$ such that $0 \le \theta_{k} \le 1$ and $\sum\theta_{k} = 1$, the mixture likelihood is

$$
\pi(\boldsymbol y \mid{\boldsymbol \alpha}, {\boldsymbol \theta}) = \sum_{k=1}^{K}\sum_{n=1}^{N}\theta_k \pi_k(y_n \mid \alpha_k).
$$


# Poisson-lognormal mixture model

Let's consider a simple example where the likelihood is given by a mixture of two Poisson-lognormal [(PoiLog)](https://mattocci27.github.io/assets/poilog.html),

$$
\pi(y_1, ..., y_N \mid \mu_1, \mu_2, \sigma_1,\sigma_2, \theta) = \sum_{n=1}^{N}\theta PoiLog(y_n \mid \mu_1, \sigma_1) + \sum_{n=1}^{N}(1-\theta) PoiLog(y_n \mid \mu_2, \sigma_2).
$$

We define two different components of Poisson-lognormal distributions, $PoiLog(\mu_1 = log(50) \simeq 3.91, \sigma_1 = 0.8)$ and $PoiLog(\mu_2 = log(400) \simeq 5.99, \sigma_2 = 0.2)$ with the mixture weight $\theta = 0.8$

```{r}
mu <- c(log(50), log(400))
sigma <- c(0.8, 0.2)
theta <- 0.8
```

```{r mix-dens}
library(cowplot)
library(ggthemes)
theme_set(theme_solarized(light=FALSE))

xx <- seq(1, 1000, by = 1)
y1 <- dnorm(log(xx), mu[1], sigma[1])
y2 <- dnorm(log(xx), mu[2], sigma[2])
y_mix <- theta * y1 + (1 - theta) * y2

line_dat <- data_frame(xx,
  comp1 = y1,
  comp2 = y2,
  comp_mix = y_mix) %>%
  gather(Model, val, 2:4) %>%
  mutate(model2 = ifelse(Model == "comp_mix",
                         "Mixed components",
                         "Each component"))

ggplot(line_dat, aes(x = xx %>% log, y = val %>% exp, col = Model)) +
  geom_line() +
  geom_point() +
  facet_wrap(~model2, scale = "free") +
  xlab("log k") +
  ylab("Density")

```

then we simulate data based on the above probability density.

```{r mix_hist}
set.seed(123)

N <- 100
z <- rbinom(N, 1, 1 - theta) + 1
log_lambda <- rnorm(N, mu[z], sigma[z])
Y <- rpois(N, exp(log_lambda))
#lambda <- rlnorm(N, mu[z] - 0.5 * sigma[z]^2, sigma[z])
#Y <- rpois(N, lambda)

hist_dat <- data_frame(Y)

p1 <- ggplot(hist_dat, aes(Y)) +
  geom_histogram(bins = 20, fill = "#268bd2", col = "black")

p2 <- ggplot(hist_dat, aes(log(Y))) +
  geom_histogram(bins = 20, fill = "#268bd2", col = "black")

plot_grid(p1, p2)


```

# Testing

## Centered parameterization

First, we fit the direct centered parameterization.

$$
\mu_1,\mu_2 \sim \mathcal{N}(0, 10)
$$

$$
\sigma_1, \sigma_2 \sim \text{Half-}Cauchy(0, 5),
$$

$$
log\lambda_{1n} \sim \mathcal{N}(\mu_1, \sigma_1)
$$

$$
log\lambda_{2n} \sim \mathcal{N}(\mu_2, \sigma_2)
$$

$$
\theta \sim Beta(2, 2)
$$


```{r}
writeLines(readLines("poilog_mix_cp.stan"))
```

The model looks straightforward but this model will suffer inefficiency in parameter sampling. Because the values of $\mu$ and $\sigma$ are highly correlated in the posterior [@Betancourt2013]. When the model is expressed this way, Stan has trouble sampling from the neck of the funnel geometry (Figure 28.1 in Stan Version 2.17.0 Manual), where $\sigma$ is small and thus $\mu$ is constrained to be near 0.


```{r, cache=TRUE, comment=NA}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

list_dat <- list(N = N, y = Y)

fit_cp <- stan(file = "poilog_mix_cp.stan",
            data = list_dat,
            iter = 2000,
            warmup = 1000,
            thin = 1,
            chains =  4,
            refresh = 500,
            control = list(adapt_delta = 0.9, max_treedepth = 20))

```

The model did not converge well. Additionally, there were divergent transitions, which indicates that the simulated Hamiltonian diverges from the true Hamiltonian. **Even a few numbers of divergent transitions indicate serious bias on the estimates.**

```{r}

print(fit_cp,
      pars = c("mu", "sigma", "theta",
               "log_lambda[1,1]",
               "log_lambda[1,2]",
               "log_lambda[1,100]",
               "log_lambda[2,1]",
               "log_lambda[2,2]",
               "log_lambda[2,100]",
               "lp__"))

```

## Non-Centered parameterization

Instead of sampling $log\lambda_{1n}$ and $log\lambda_{2n}$ directly, the model can be converted to the following more efficient form,

$$
\mu_1,\mu_2 \sim \mathcal{N}(0, 10)
$$

$$
\sigma_1, \sigma_2 \sim \text{Half-}Cauchy(0, 5),
$$

$$
log\tilde\lambda_{1n},log\tilde\lambda_{2n}  \sim \mathcal{N}(0,1)
$$

$$
log\lambda_{1n} = \mu_1 + log\tilde\lambda_{1n} \times \sigma_1
$$

$$
log\lambda_{2n} = \mu_2 + log\tilde\lambda_{2n} \times \sigma_2
$$

$$
\theta \sim Beta(2, 2)
$$


```{r}
writeLines(readLines("poilog_mix_ncp.stan"))
```

This non-centered parameterization exchanges a direct dependence between ($\mu$, $\sigma$) and $log\lambda$ for a dependence between ($\mu$, $\sigma$) and $y_i$, which would reduce the dependence between $\lambda$, $\mu$, and $\sigma$.

```{r, cache=TRUE, comment=NA}

fit_ncp <- stan(file = "poilog_mix_ncp.stan",
            data = list_dat,
            iter = 2000,
            warmup = 1000,
            thin = 1,
            chains =  4,
            refresh = 500,
            control = list(adapt_delta = 0.9, max_treedepth = 20))

```

Now we don't have divergent transitions. The resulting estimates well recovered the original values ($\boldsymbol \mu$, $\boldsymbol \sigma$ and $\theta$) `r emo::ji("beer")`. Note that the above model will not work very well when the two components are relatively close ([see detailed explanation](https://betanalpha.github.io/assets/case_studies/identifying_mixture_models.html)).


```{r}
print(fit_ncp,
      pars = c("mu", "sigma", "theta",
               "log_lambda[1,1]",
               "log_lambda[1,2]",
               "log_lambda[1,100]",
               "log_lambda[2,1]",
               "log_lambda[2,2]",
               "log_lambda[2,100]",
               "lp__"))
```

```{r, eval = F, echo = F}
pairs(fit_ncp, pars=c("mu", "sigma"))
pairs(fit_cp, pars=c("mu", "sigma"))
```

# Computing Environment

```{r}
writeLines(readLines(file.path(Sys.getenv("HOME"), ".R/Makevars")))
devtools::session_info("rstan")
```

# Related

- [Mixture model (wiki)](https://en.wikipedia.org/wiki/Mixture_model)

- [Fitting a Poisson-lognormal distribution in Stan (rstan)](https://mattocci27.github.io/assets/poilog.html)

- [Diagnosing Biased Inference with Divergences](https://betanalpha.github.io/assets/case_studies/divergences_and_bias.html#3_a_non-centered_eight_schools_implementation)

- [Identifying Bayesian Mixture Models
](https://betanalpha.github.io/assets/case_studies/identifying_mixture_models.html)

- [Poisson-LogSkewNormal mixture + partial pooling](https://discourse.mc-stan.org/t/poisson-logskewnormal-mixture-partial-pooling/2189)

# References
